---
title: "Semiparametric regression - Homework 3"
author: "Klaudia Weigel"
output: 
  pdf_document: 
    fig_caption: yes
    highlight: tango
    number_sections: yes

header_includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{amsthm}
  - \usepackage{listings}
  - \theoremstyle{definition}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1

We define the following functions on $[0,1]$:
$$
T_1(x) = 1, \qquad T_2(x) = x, \qquad T_3(x) = (x-\frac{1}{2})_+ = \max(x-\frac{1}{2}, 0),
$$

$$
B_1(x) = (1-2x)_{+}, \qquad B_2(x) = 1 - |2x-1|, \qquad B_3(x) = (2x - 1)_+,
$$

where $x \in \mathbb{R}$.

## (a)

We will plot the previously defined functions.
```{r}
ng <- 101
xg <- seq(0, 1, length=ng)
T1g <- rep(1, ng)
T2g <- xg
T3g <- (xg - .5)*(xg - .5>0)
B1g <- (1 - 2*xg)*(1 - 2*xg>0)
B2g <- 1 - abs(2*xg - 1)
B3g <- 2*T3g
par(mfrow=c(2,1), mar=c(2, 4, 2, 2))
plot(0, type = "n", xlim=c(0,1), ylim=c(0,1), xlab="x", ylab="") #, bty="1")
lines(xg, T1g, col=1)
lines(xg, T2g, col=2)
lines(xg, T3g, col=3)
text(0.1, 0.8, expression(T[1]), col=1)
text(0.4, 0.5, expression(T[2]), col=2)
text(0.8, 0.2, expression(T[3]), col=3)
plot (0, type = "n", xlim=c(0,1), ylim=c(0,1), xlab="x", ylab="") #, bty="1")
lines(xg, B1g, col=4)
lines(xg, B2g, col=5)
lines(xg, B3g, col=6)
text(0.1, 0.9, expression(B[1]), col=4)
text(0.4, 0.9, expression(B[2]), col=5)
text(0.6, 0.6, expression(B[3]), col=6)
```


## (b)

We will now find expressions for $B_1, B_2, B_3$ in terms of $T_1, T_2$ and $T_3$.

$$
\begin{aligned}
B_1(x) &= \max(1-2x,0) = 1 - 2x - (1-2x)\mathbb{I}_{\{x>1/2\}} = 1 - 2x + 2(x - \frac{1}{2})\mathbb{I}_{\{x>1/2\}} \\
&= 1 - 2x + 2 \max(x-\frac{1}{2} , 0) = T_1(x) - 2T_2(x) + 2T_3(x).
\end{aligned}
$$

$$
\begin{aligned}
B_2(x) &= 1 - |2x - 1| = 1 + 2x - 1 - 2(2x-1)\mathbb{I}_{\{x > 1/2\}} = 2x - 4(x-\frac{1}{2})\mathbb{I}_{\{x > 1/2\}} \\
&= 2x  - 4 \max(x-\frac{1}{2} , 0) = 2T_2(x) - 4T_3(x),
\end{aligned}
$$

$$
B_3(x) = \max(2x-1, 0) = 2\max(x - \frac{1}{2}, 0) = 2T_3(x).
$$
Next we'll check what $B_1 + B_2 + B_3$ is:

$$
B_1 + B_2 + B_3 = T_1 - 2T_2 + 2T_3 + 2T_2 - 4T_3 + 2T_3 = T_1 =1.
$$

We can check the results with a plot.
```{r}
par(mfrow=c(2,1), mar=c(2, 4, 2, 2))
plot (0, type = "n", xlim=c(0,1), ylim=c(0,1), xlab="x", ylab="") #, bty="1")
lines(xg, T1g - 2*T2g + 2*T3g, col=4)
lines(xg, 2*T2g - 4*T3g, col=5)
lines(xg, 2*T3g, col=6)
text(0.1, 0.6, expression(T[1] - 2*T[2] + 2*T[3]), col=4)
text(0.35, 0.9, expression(2*T[2] - 4*T[3]), col=5)
text(0.6, 0.4, expression(2*T[3]), col=6)
plot (0, type = "n", xlim=c(0,1), ylim=c(0,1), xlab="x", ylab="") #, bty="1")
lines(xg, B1g + B2g + B3g, col=4)
text(0.35, 0.9, expression(B[1] + B[2] + B[3]), col=4)
```



## (c)

We will obtain $3 \times 3$ matrix $L_{TB}$ such that 
$$
[B_1(x)\quad B_2(x)\quad B_3(x)] = [T_1(x)\quad T_2(x)\quad T_3(x)]L_{TB}
$$
for any $x \in [0,1]$.

We have:
$$
[B_1 \quad B_2 \quad B_3] =  [T_1 - 2T_2 + 2T_3 \quad 2T_2 - 4T_3 \quad 2T_3] = [T_1\quad T_2 \quad T_3] \begin{bmatrix} 1 & 0 & 0\\-2 & 2 & 0 \\ 2 & -4 & 2  \end{bmatrix}.
$$

Therefore 
$$
L_{TB} = \begin{bmatrix} 1 & 0 & 0\\-2 & 2 & 0 \\ 2 & -4 & 2  \end{bmatrix}.
$$

## (d)

```{r}
LTB <- matrix(c(1,0,0,  -2,2,0,  2,-4,2), byrow = TRUE, nrow = 3)
det(LTB)
```

The determinant of $L_{TB}$  is equal to 4, therefore the matrix is invertible.

## (e)

Let
$$
X_T = \begin{bmatrix} T_1(x_1) & T_2(x_1) & T_3(x_1) \\ \vdots & \vdots & \vdots \\ T_1(x_n) & T_2(x_n) & T_3(x_n) \end{bmatrix}, \qquad
X_T = \begin{bmatrix} B_1(x_1) & B_2(x_1) & B_3(x_1) \\ \vdots & \vdots & \vdots \\ B_1(x_n) & B_2(x_n) & B_3(x_n) \end{bmatrix}
$$
be design matrices. We will check if the fit obtained using matrix $X_T$ is the same as that with $X_B$.


```{r}
set.seed(1)
n <- 100
x <- sort(runif(n))
y <- cos(2*pi*x) + 0.2*rnorm(n)
plot(x,y,col='dodgerblue', bty='l')
XT <- cbind(rep(1,n), x, (x - .5)*(x - .5 > 0))
XB <- cbind((1 - 2*x)*(1 - 2*x > 0), 1 - abs(2*x - 1), (2*x - 1)*(2*x - 1 > 0))
fitT <- lm(y~XT-1)
fitB <- lm(y~XB-1)
lines(x, fitted(fitT), col = 'orange', lwd = 6)
lines(x, fitted(fitB), col = 'darkgreen', lwd = 2)

```

We see that both fitted lines are the same.


